Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-28 paper.tex
ENH:  Preparing for merge.

\selectlanguage{eng}
\begin{document}
\onecolumn
\firstpage{1}

@@

\begin{abstract}
Diffeomorphic maps are central to image registration due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information (i.e. regularization) is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, a free-form deformation methodological variant, the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}, can be viewed as an alternative smoothing approach in which explicit regularization is achieved through 
fast B-spline appromixation.
This characterization can be used to provide alternative B-spline ``flavored'' diffeomorphic image registration solutions with several advantages which we describe in this work.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
open science, best practices, reproducibility, software }
\end{abstract}

@@

Free-form deformation (FFD) image registration, characterized by
regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity, good performance,
desirable properties, and several available implementations.  Current
research was preceded by related work for geometric modeling
\citep{sederberg1986} and originated with such important contributions
as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued
development within this early spline-based paradigm produced
additional innovations such as integrated similarity metrics
\citep[e.g.][]{mattes2003}, additional transformation constraints
\citep[e.g.][]{rohlfing2003}, and notable open source implementations
\citep[e.g.][]{ibanez2005,klein2010,shackleford2010}.

@@

It is interesting to note that within this context, traditional FFD
algorithms can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space of
regularized fields using Gaussian convolution, gradient fields are
projected to a smooth space characterized by the B-spline basis
functions.  This analogy was hinted at in our earlier work
\citep{tustison2009} where we showed that fitting the update field to
a B-spline object using a fast approximation routine
\citep{tustison2006} is equivalent to a preconditioning of the
standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect
induced by the ill-conditioned nature of the traditional
gradient-based FFD formulation.  We denote this new FFD variant as
directly manipulated free-form deformation (DMFFD) and, as part of the
ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

@@

Although presenting a rigorous framework for image registration solutions
with desirable properties, the complexity of these diffeomorphic methodologies
requires substantial computational resources.  For typical 3-D neuroimaging
applications, the corresponding solutions require numerical integration over
and storage of 4-D velocity fields at each iteration which is limiting for
many common, contemporary processing infrastructures.

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is possible since the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Furthermore, the forward and inverse
mappings are guaranteed to be consistent within the discrete domain i.e. $\|
\phi_i^{-1}(\phi_i) - \mathbf{Id} \|^2 < \epsilon$.   
Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Similarly, in the case of $d$-dimensional time-parameterized 
diffeomorphic image registration, 
the time-dependent velocity field can be represented
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field.  The preconditioned gradient analog of
Equation (\ref{eq:dmffd}) for updating the time-varying velocity field control point lattice is 
\begin{align}
\label{eq:tvdmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c)
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which takes into account the temporal locations of the
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
\footnote{
Additionally, in \cite{tustison2006} it was shown that one could associate each
a sample, $\left( \frac{\partial \Pi}{\partial \mathbf{x}}\right)_c$ with a confidence
weighting.  Thus, in order to enforce stationary boundaries for all described
DMFFD-based diffeomorphic algorithms, we
assign image boundary metric gradients a value of zero with a corresponding
large confidence value.
}

@@

@@

B-spline SyN produced slightly greater Dice values than the original SyN.
Interestingly it was shown that the range of log Jacobian values tend to be
significantly higher for the former than the latter.  Although possible 
explanations include the approximation-versus-convolution distinction mentioned
earlier, further exploration is planned.

>>>>

%\selectlanguage{eng}
\begin{document}
\onecolumn
\firstpage{1}

@@

\begin{abstract}
Diffeomorphic maps are central to image registration due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility 
taking the form of physics-based constraints or through some approximation thereof, e.g. 
Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In 
the context of the original Demons' framework, the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}
can be viewed as a smoothing alternative  in which explicit regularization is achieved through 
fast B-spline approximation.
This characterization can be used to provide B-spline ``flavored'' diffeomorphic image 
registration solutions with several advantages.  Implementation 
is open source and available through the Insight Toolkit and our Advanced Normalization Tools
(ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm 
\citep{avants2008}, implemented within the same framework, and its B-spline analog is 
performed using open labeled brain data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Free-form deformation (FFD) image registration, characterized by
regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity, good performance,
desirable properties, and several available implementations.  Current
research was preceded by related work for geometric modeling
\citep{sederberg1986} and originated with such important contributions
as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued
development within this early spline-based paradigm produced
additional innovations such as integrated similarity metrics
\citep[e.g.][]{mattes2003}, additional transformation constraints
\citep[e.g.][]{rohlfing2003}, and notable open source implementations
\citep[e.g.][]{ibanez2005,klein2010,modat2010,shackleford2010}.

@@

It is interesting to note that within this context, traditional FFD
algorithms can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space of
regularized fields using Gaussian convolution, gradient fields are
projected to a smooth space characterized by the B-spline basis
functions.  This analogy was hinted at in our earlier work
\citep{tustison2009} where we showed that fitting the update field to
a B-spline object using a fast approximation routine
\citep{tustison2006} is equivalent to a preconditioning of the
standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect
induced by the ill-conditioned nature of the traditional
gradient-based FFD formulation.  We denoted this new FFD variant as
directly manipulated free-form deformation (DMFFD) and, as part of the
ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

@@

Although presenting a rigorous framework for image registration solutions
with desirable properties, the complexity of these diffeomorphic methodologies
requires substantial computational resources.  For typical 3-D neuroimaging
applications, the corresponding solutions require numerical integration over
and storage of 4-D velocity fields at each iteration which is limiting for
many common processing infrastructures.

Therefore, in addition to the full-scale
SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in differents applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is possible since the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Furthermore, the forward and inverse
mappings are guaranteed to be consistent within the discrete domain i.e. $\|
\phi_i^{-1}(\phi_i) - \mathbf{Id} \|^2 < \epsilon$.   
Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Similarly, in the case of $d$-dimensional time-parameterized 
diffeomorphic image registration, 
the time-dependent velocity field can be represented
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field.  The preconditioned gradient analog of
Equation (\ref{eq:dmffd}) for updating the time-varying velocity field control point lattice is 
\begin{align}
\label{eq:tvdmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c)
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which takes into account the temporal locations of the
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
\footnote{
Additionally, in \cite{tustison2006} it was shown that one could associate each
a sample, $\left( \frac{\partial \Pi}{\partial \mathbf{x}}\right)_c$ with a confidence
weighting.  Thus, in order to enforce stationary boundaries for all described
DMFFD-based diffeomorphic algorithms, we
assign to the image boundary metric gradients the null vector with a corresponding
large confidence value.
}

@@

B-spline SyN produced slightly greater Dice values than the original SyN. 
Although actual differences are relatively small, they are statistically 
significant. By implementing both algorithms in the the same code base, 
we are not only able to eliminate non-regularization components of the 
registration but we are also able to eliminate implementation differences. 
Thus, performance disparity can be isolated to smoothing choice. However, 
even with this restricted focus there are various reasons for the evaluation 
outcome. These include the approximation-versus-convolution distinction 
mentioned earlier for the two regularization approaches (which could also 
explain the reason why the range of log Jacobian values tend to be significantly 
higher for B-spline SyN). Also, the fact that the regularization for 
B-spline SyN is theoretically continuous whereas the truncated Gaussian 
convolution is only a discrete approximation could be a potential factor. 
Exploration of other possibilities are planned for future work.

@@
